Assessment of forestry road network projection methods using the “roads” package in R

### 2022-10-14 analysis/paper/paper.Rmd
paper edits

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers. Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope [@governmentofbritishcolumbia2021a]. Major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0 (Table 1). The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs used to build the cost raster.

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)). Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the cost of building a road at any point between the two. The path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is determined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed. In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method. This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks. Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.

@@

```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks. The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP). Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density. Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the cost of road building by the grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods. We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced pointd or random points, with point densities of 1 and 10 pts/km^2^ (Figure \@ref(fig:fig2)). The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings. For the Hardy QGIS plugin we used the 100 m resolution cost raster. We kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^. This resulted in a total of 15 road projection variants to be compared (Table 2).

@@

In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road. 4) Road disturbance footprint, 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks; The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes]. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004]. Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package. Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads. All road characteristics were stored in raster format.

The aspatial accuracy of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and at the cutblock level, where metric rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.

The projections were assessed for their spatial accuracy in two different ways. First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to highlight discrepancies between the networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy. Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed. All pixels in the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed. These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarily or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest roads were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).

>>>>

Our road network data was obtained by combining the digital road atlas (DRA) [@databc2021] and forest tenure road [@britishcolumbiaministryofforestslandsnaturalresourceoperationsandruraldevelopmentflnrord2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990, all roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@governmentofbritishcolumbia2021b]. Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@governmentofbritishcolumbia2021b]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

To spatially define the cost of building and traveling on a forestry road we developed a simple cost raster based on the slope and major barriers. Following the Interior Appraisal Manual of BC, the base cost per kilometer of road was a linear function of slope (Table 1) [@governmentofbritishcolumbia2021a]. Major water bodies were considered barriers and assigned a very high cost. Lastly, existing roads were assigned a cost of 0. The cost raster produced consists of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road construction costs used to build the cost raster.

@@

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the cost of traversing the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, dynamic least cost paths and minimum spanning trees (Figure \@ref(fig:fig2)). Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm [-@dijkstra1959] to determine the optimal path between a cutblock and the existing road based on the road building cost between the two. The least cost path to each cutblock is determined independently of the other cutblocks. The dynamic least cost paths (DLCP) method uses the same LCP algorithm but evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. The sequence is determined by a heuristic such as the distance to the existing road. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm [-@kruskal1956]. (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the DLCP and MST approaches which are better aligned with how road networks are designed (Figure \@ref(fig:fig2)). In addition to the projection methods included in the `roads` package, we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>), hereafter referred to as the Hardy QGIS method. This plugin uses a similar DLCP algorithm to the `roads` package and uses the distance to the closest roads to order the cutblocks. Finally, in some cases the impact of forestry on wildlife is projected without considering the road development that will accompany it, we refer to this as the cutblocks only projection method and compare it to projections that do consider roads.

@@

```{r fig2, fig.cap="The roads package includes several different algorithms for building road networks, as well as methods for selecting landing locations within forest harvest cublocks. The algorithms include minimum spanning trees (MST), dynamic least cost paths (DLCP) and least cost paths (LCP). Landing locations can be selected as the centroid of the cutblock, or by taking a random or regularly spaced sample of points inside the cutblock at a given density. Target cutblocks are shown by grey lines, existing roads by black lines, projected roads by yellow lines, observed roads by red lines, and the 100 m resolution cost raster by the grey shading."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

We used the development of roads in the Revelstoke TSA from 1990 to the present as a case study to compare the MST, DCLP and Hardy QGIS road projection methods. We also explored the effects of several variations of the input parameters by repeating the MST projections for each variant. We compared two resolutions of the cost surface by aggregating the 100m resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points within cutblocks by using the polygon centroid, regularly spaced points or random points. Regular and random  point densities of 1 and 10 pts/km^2^ were tested (Figure \@ref(fig:fig2)). The MST projection method was used for all the variations of parameters described above and an additional projection was run using the DLCP method with regularly spaced, high density landings. For the Hardy QGIS plugin we used the 100 m resolution cost raster. We kept all the default settings except for the skidding distance which was set to 200m. Note that a skidding distance of 200 m corresponds to \~ 9 pts/km^2^. This resulted in a total of 13 road projection variants to be compared (Table 2).

@@

In order to assess the performance of the projections, the following characteristics of the projected and observed road networks were calculated and compared: 1) Road presence, whether or not roads are present in each pixel; 2) Road density, density of roads in meters per hectare (m/ha); 3) Distance to nearest road, distance from each raster cell to the nearest road. 4) Road disturbance footprint, a 1 km buffer around roads [@ibisch2016a]; 5) Forestry disturbance footprint, a 500 m buffer around roads and cutblocks. The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou [@johnson2020], and was calculated using the `caribouMetrics` R package [@hughes]. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife [@beazley2004]. Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package. Distance to nearest road was calculated with a fast moving window method using the getDistFromSource function of the `roads` package. Only the road and forestry disturbance footprints were calculated for the cutblocks only method since the other characteristics are not applicable with out roads. All road characteristics were stored in raster format.

The aspatial performance of the projections was assessed by calculating the mean values of each road network characteristic and comparing the projected and observed values. This was done across the TSA and at the cutblock level, where road network characteristic rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.

The projections were assessed for their spatial performance in two different ways. First, difference maps were created by subtracting the projected distance to the nearest road from the observed in order to highlight discrepancies between the road networks. The observed and projected road networks were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS. Secondly, the performance of each projection for the three binary characteristics (road presence, road disturbance footprint, and forestry disturbance footprint) was assessed. All pixels in the projected rasters fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed. These categories were summarized for each projection and characteristic and used to calculate the performance metrics sensitivity, precision and F-measure [@lee2021]. Sensitivity, also known as recall, gives the proportion of the true positives that were projected, while precision gives the proportion of positives projected that are true. Sensitivity will be lower when observed roads are missed and precision will be lower when roads are projected that were not observed. F-measure attempts to balance sensitivity and precision to give an overall score.

@@

For each projection method variant the aspatial performance varied across the 5 road network characteristics and between the within cutblock and overall scales (Figure \@ref(fig:fig3)). The buffering based characteristics, the road and forestry disturbance footprints, performed best within cutblocks, while the other characteristics, presence, density and distance to nearest road, performed similarly or better across the whole landscape. Across the whole landscape the road and forestry disturbance footprints and distance to nearest road characteristics were the closest to the observed mean, while the road presence and density characteristics were underestimated. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (\@ref(fig:figA1)).
